The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-21 manuscript/rebuttal.Rmd

> 3. Page 3, Line 43: "enshrined by code" (this language is awkward)

RESPONSE: The passage has been amended for clarity:

> "... it needs to be formalised in code and ... "

> 4. Page 3, Lines 45-47: What about tasks that cannot be automated? I see that this topic is addressed
later. But this part implies that everything can be automated.

RESPONSE: The section has been amended to acknowledge that not all steps can be automated.

> To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
> possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.

> 5. Page 3, Line 52: It says that spreadsheets are "overused and misused." This is subjective and not
> backed by evidence, other than the well-known examples of gene symbols being formatted as dates.

RESPONSE: The passage in question has been replaced, with reference to evidence from a large
survey that found that ~69% of university academics used MS Excel in their research activities.

> Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
> researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

> 6. Page 3, Line 56: It is not necessarily true that analyses performed using web tools are not
> reproducible. Although rare, some web tools facilitate reproducibility by providing code or
> configuration files and/or allowing the apps to be executed locally.

### 2023-08-21 manuscript/5pillars.Rmd
working on reviewer comments

For a bioinformatics project to be reproducible, it needs to be formalised in code and include all
instructions, from inspecting the raw data to generating the outputs that form the conclusions of the
study.
This automated process removes the need for any manual steps, which are time-consuming and
prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
somewhat overused and misused in the life sciences, leading to problems like inadvertent conversion
of gene names to dates in databases and supplementary Excel files [@Zeeberg2004-js;
@Ziemann2016-pv; @Abeysooriya2021-uz].
While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.

>>>>

To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
Automated processes remove the need for manual steps, which are time-consuming and
prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
leading to problems like inadvertent conversion of gene names to dates in databases and supplementary
Excel files [@Zeeberg2004-js; @Ziemann2016-pv; @Abeysooriya2021-uz].
Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.
